% \documentclass[sigplan,screen,9pt]{acmart}
\documentclass[sigplan,screen,authorversion,9pt]{acmart}
\usepackage{booktabs}
\usepackage{subcaption}

% Haskell code snippets and useful shortcuts
\usepackage{minted}
\setminted[haskell]{escapeinside=@@}
\newcommand{\hs}{\mintinline{haskell}}
\newcommand{\teq}{\smaller $\sim$}
\newcommand{\ghci}{$\lambda$>}
\newcommand{\defeq}{\stackrel{\text{def}}{=}}
\newcommand{\std}[1]{{\color[rgb]{0,0.3,0} #1}}
\newcommand{\blk}[1]{{\color[rgb]{0,0,0} #1}}

\makeatletter

%%% The following is specific to Haskell'17 and the paper
%%% 'Algebraic Graphs with Class (Functional Pearl)'
%%% by Andrey Mokhov.
%%%
\setcopyright{acmlicensed}
\acmPrice{15.00}
\acmDOI{10.1145/3122955.3122956}
\acmYear{2017}
\copyrightyear{2017}
\acmISBN{978-1-4503-5182-9/17/09}
\acmConference[Haskell'17]{10th ACM SIGPLAN International Haskell Symposium}{September 7-8, 2017}{Oxford, UK}

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}

\begin{document}
\title{Algebraic Graphs with Class (Functional Pearl)}

\author{Andrey Mokhov}
\affiliation{
  \institution{Newcastle University, United Kingdom}
}

\begin{abstract}
% \vspace{-0.5mm}
The paper presents a minimalistic and elegant approach to working
with graphs in Haskell. It is built on a rigorous
mathematical foundation --- an algebra of graphs --- that allows us to apply
equational reasoning for proving the correctness of graph transformation
algorithms. Algebraic graphs let us avoid partial functions typically
caused by `malformed graphs' that contain an edge referring to
a non-existent vertex. This helps to liberate APIs of existing graph libraries
from partial functions.

The algebra of graphs can represent directed, undirected, reflexive
and transitive graphs, as well as hypergraphs, by appropriately choosing
the set of underlying axioms. The flexibility of the approach is
demonstrated by developing a library for constructing
and transforming polymorphic graphs.%\vspace{-1mm}
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002950</concept_id>
<concept_desc>Mathematics of computing</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Mathematics of computing}
% \keywords{\vspace{-2mm}Haskell, algebra, graph theory}
\keywords{Haskell, algebra, graph theory}

\maketitle

\input{intro}
\input{core}
\input{algebra}
\input{graphs}
\input{library}
\input{related-work}
\input{discussion}

%% Acknowledgments
\begin{acks}
  I would like to thank Arseniy Alekseyev and Neil Mitchell
  for numerous discussions on algebraic graphs, their advice,
  criticism and encouragement. Without their help this work would
  have likely remained an unfinished toy project.
  Simon Peyton Jones, Brent Yorgey, Danil Sokolov, Ben Lippmeier, Ulan Degenbaev
  and several anonymous reviewers provided constructive feedback on an
  earlier draft of this paper, helping to substantially improve it. Last but
  not least, I'm very grateful to Victor Khomenko for his contribution to the
  algebra of parameterised graphs that forms the mathematical foundation of this work.
\end{acks}

\bibliography{publications}

\end{document}
